Impact Analysis: Health and Economic Cost of United States Storm Events
==================================================================================

Synopsis
========

The following report attempts to quantify the human and economic costs of storm events 
across the United States.  The US National Oceanic and Atmospheric Administration (NOAA)
database houses storm event data going back to the 1950's. This database houses data--
including counts of fatalities, injuries, property damage and crop damage--on significant
US storm events. The following report contains the r code necessary to process and plot
the data, ultimately answering which types of storm events are most harmful to human health
and which types inflict the greatest economic damage.

Data Processing
==================
#### Load Packages
```{r, load_packages}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(gridExtra))
suppressMessages(library(outliers))
```
#### Load Data
First, download and load the raw data into R. To make manipulation easier,I transform 
column names to lowercase and put the date variable into a suitable date format.
'raw_copy' functions as a clean original data set which is not to be manipulated.
```{r, load_data, cache = TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists('repdata-data-StormData.csv.bz2')){
  download.file(url, destfile = 'repdata-data-StormData.csv.bz2')
}
data <- read.csv("repdata-data-StormData.csv.bz2")
names(data) <- tolower(names(data))
data$bgn_date <- as.Date(data$bgn_date, "%m/%d/%Y")
raw_copy <- data
```
#### Format Property and Crop Damage Estimates
Property and crop damage estimates are currently recorded as a numeric value and
placeholder value--K, M, B.  I combine the two for a more interpretable value.

```{r, dollar_value_formatting}
data$property_dmg <- ifelse(data$propdmgexp == "K", data$propdmg * 1e3,
                                 ifelse(data$propdmgexp == "M", data$propdmg * 1e6,
                                        ifelse(data$propdmgexp == "B", data$propdmg * 1e9,
                                               data$propdmg * 0)))

data$crop_dmg <- ifelse(data$cropdmgexp == "K", data$cropdmg * 1e3,
                                 ifelse(data$cropdmgexp == "M", data$cropdmg * 1e6,
                                        ifelse(data$cropdmgexp == "B", data$cropdmg * 1e9,
                                               data$cropdmg * 0)))
```
#### Checking For Outliers
An initial check for outliers is always useful.  In this case, with over 900,000 records,
it is very possible we have data entry errors which could mess with the analysis.
Using the outlier package I identified the values furthest from the mean in each of 
our four categories. I use our raw data set so I can examine any suspicious entries.
```{r}
outliers <- data %>%
  filter(property_dmg == outlier(data$property_dmg)|crop_dmg == outlier(data$crop_dmg)|
           fatalities == outlier(data$fatalities)|injuries == outlier(data$injuries))
as.character(outliers$remarks[5]) #Property Damage outlier statement
```
Examining the records, all the outliers except property damage appear to be legitimate.
Reading the remark for property damage, it appears the value was supposed to be 
~76 million.  The value 115 BILLION is over 1500 times that amount.  It appears
the propertydmgexp was incorrectly entered as a 'B' instead of 'M'.  I exclude this 
value in the rest of the analysis.
```{r}
data <- data%>%
  filter(property_dmg != outlier(data$property_dmg))
```

Now subset the data set to only contain values I need for answering my questions--
fatalities, injuries, property_dmg, crop_dmg, evtype and bgn_date.
```{r, data_subset}
data <- data %>%
  select(bgn_date, evtype, fatalities, injuries, property_dmg, crop_dmg)
data$evtype <- tolower(data$evtype)
head(data, n = 2)
```
#### Examining Event Classification Scheme

The current NOAA storm classification scheme consists of 48 distinct event types. However, 
this scheme only went into effect in 1996. Whether to examine the entire data set or
limit my exploration to post-1995 data is a key decision in my analysis.

Limiting to post-1995 data may remove rare, but of significant impact, storm events.  
Examining the full data set may skew certain aspects of the data.  For instance, 
the calculation of damage estimates could vary over time or storm categorization
may vary over time.

A good starting place is to look at the proportion of the relevant damage taking place
before 1996.
```{r}
pre_1996 <- data %>%
  filter(bgn_date < '1996-01-01')
post_1996 <- data %>%
  filter(bgn_date >= '1996-01-01')
pre_death <- round(sum(pre_1996$fatalities)/sum(data$fatalities) * 100, digits = 1)
pre_injury <- round(sum(pre_1996$injuries)/sum(data$injuries) * 100, digits = 1)
pre_prop <- round(sum(pre_1996$property_dmg)/sum(data$property_dmg) * 100, digits =1)
pre_crop <- round(sum(pre_1996$crop_dmg)/sum(data$crop_dmg) * 100, digits = 1)
```

`r pre_death`% and `r pre_injury`% of storm deaths and injuries, respectively, took
place prior to 1996. `r pre_prop`% and `r pre_crop`% of property and crop damage, 
respectively, took place prior to 1996.

If you look at the 5 largest events for the four areas of interest it becomes
apparent that while the majority of economic damage is attributed to storms post-1995,
a huge portion of the human cost of storms is accounted for by storms prior to 1996.
```{r, damage_breakdown, results = 'hold'}
options(scipen = 20)
data %>%
  select(bgn_date, evtype, fatalities)%>%
  arrange(desc(fatalities))%>%
  top_n(5, fatalities)
data %>%
  select(bgn_date, evtype, injuries)%>%
  arrange(desc(injuries))%>%
  top_n(5, injuries)
data %>%
  select(bgn_date, evtype, property_dmg)%>%
  arrange(desc(property_dmg))%>%
  top_n(5, property_dmg)
data %>%
  select(bgn_date, evtype, crop_dmg)%>%
  arrange(desc(crop_dmg))%>%
  top_n(5, crop_dmg)
```

Maybe early warning systems have improved post-1995 and there are fewer injuries or 
deaths due to storms.  Maybe the estimation procedure for financial damage has changed
post-1995.  It is pure speculation without more information.  Regardless of the reason,
I believe it is necessary to look at entire data set.

#### Event Type Preprocessing
Next, I have to deal with the huge number of event types present in our data set.
```{r, results = 'hide', echo = FALSE}
num_evtype <- length(unique(data$evtype))
```
With `r num_evtype` unique event classifications, plotting would be unweldy. More 
importantly, many of these types are data entry errors or different ways of entering 
identical storm events.

My initial approach was to try out R's stringdist package and cluster evtypes based on
string similarity.  Despite a fair amount of tinkering, I could never get a satisfactory
set of clusters.  Manual classification is more work, but seemed to be my best option.

If you look at the top 20 deaths by event type, you immediately see several events 
that probably belong to one category.
```{r, events_by_death}
data %>%
  group_by(evtype)%>%
  summarise(count = n(), deaths = sum(fatalities), injuries = sum(injuries), 
            prop_dmg = sum(property_dmg), crop_dmg = sum(crop_dmg)) %>%
  arrange(desc(deaths)) %>%
  top_n(20, deaths)
```
'tstm wind' and 'thunderstorm wind' or 'extreme cold' and 'extreme cold/wind chill' 
are surely are the same thing. There are many other examples of this as you sift through 
```{r, cache = TRUE}
unique_events <- data %>%
  group_by(evtype)%>%
  summarise(count = n(), deaths = sum(fatalities), injuries = sum(injuries), 
            prop_dmg = sum(property_dmg), crop_dmg = sum(crop_dmg)) %>%
  filter(deaths > 0 | injuries > 0 | prop_dmg > 0 | crop_dmg > 0)
```
There are `r nrow(unique_events)` unique storm event names where either deaths, injuries, 
property damage or crop damage are greater than 0.  I am manually reclassifying the 
messy event records into as close to the 48 official categories as I can get.

```{r, event_reclassification, cache = TRUE}
unique_events$evtype <- ifelse(grepl("torn", unique_events$evtype) == TRUE, "tornado", 
                               unique_events$evtype)
unique_events$evtype <- ifelse(grepl("avalanc", unique_events$evtype) == TRUE, 
                               "avalanche", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("blizzard", unique_events$evtype) == TRUE,
                               "blizzard", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("coastal|cstl", unique_events$evtype) == TRUE,
                               "coastal flood", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("cold|low temp", unique_events$evtype) == TRUE &
                        grepl("extreme", unique_events$evtype) == FALSE,
                        "cold/wind chill", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("drought", unique_events$evtype) == TRUE, 
                               "drought", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("dust devil", unique_events$evtype) == TRUE,
                               "dust devil", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("dust storm", unique_events$evtype) == TRUE, 
                               "dust storm", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("excessive heat|extreme heat|record heat", 
                                     unique_events$evtype) == TRUE, "excessive heat", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("extreme cold|extreme wind|hypothermia",
                                     unique_events$evtype) == TRUE,
                               "extreme cold/wind chill", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("flash", unique_events$evtype) == TRUE,
                               "flash flood", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("flood|fld", unique_events$evtype) == TRUE & 
                                 grepl("coastal flood|flash flood|lakeshore flood",
                                       unique_events$evtype) == FALSE, "flood", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("freeze|frost|freezing", unique_events$evtype) == TRUE & 
                        grepl("freezing fog", unique_events$evtype) == FALSE,
                        "frost/freeze", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("funnel", unique_events$evtype) == TRUE,
                               "funnel cloud", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("hail", unique_events$evtype) == TRUE &
                        grepl("marine", unique_events$evtype) == FALSE,
                        "hail", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("heat|warm", unique_events$evtype) == TRUE & 
                        grepl("excessive heat", unique_events$evtype) == FALSE, 
                        "heat", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("heavy|precip", unique_events$evtype) == TRUE &
                        grepl("snow|surf|swells|seas", unique_events$evtype) == FALSE,
                        "heavy rain", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("heavy|excessive snow|record snow", unique_events$evtype) == TRUE &
                        grepl("rain|swells|seas", unique_events$evtype) == FALSE,
                        "heavy snow", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("hurricane|typhoon", unique_events$evtype) == TRUE,
                               "hurricane (typhoon)", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("high surf|seas|swells|surf", unique_events$evtype) == TRUE, 
                               "high surf", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("ice|icy", unique_events$evtype) == TRUE,
                               "ice storm", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("lake effect|lake-effect", unique_events$evtype) == TRUE,
                               "lake-effect snow", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("lightning", unique_events$evtype) == TRUE,
                               "lightning", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("marine thunderstorm wind|marine tstm wind",
                                     unique_events$evtype) == TRUE,
                               "marine thunderstorm wind",unique_events$evtype)
unique_events$evtype <- ifelse(grepl("rip current", unique_events$evtype) == TRUE,
                               "rip current", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("sleet", unique_events$evtype) == TRUE,
                               "sleet", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("storm surge", unique_events$evtype) == TRUE,
                               "storm surge/tide", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("strong wind", unique_events$evtype) == TRUE &
                        grepl("marine|ice", unique_events$evtype) == FALSE,
                        "strong wind", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("thun|tstm|microburst", unique_events$evtype) == TRUE &
                         grepl("marine|snow|non", unique_events$evtype) == FALSE, 
                         "thunderstorm wind", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("tropical storm", unique_events$evtype) == TRUE,
                               "tropical storm", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("waterspout", unique_events$evtype) == TRUE,
                               "waterspout", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("fire", unique_events$evtype) == TRUE,
                               "wildfire", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("winter storm", unique_events$evtype) == TRUE,
                               "winter storm", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("winter weather|wintry|snow|glaze", unique_events$evtype) == TRUE &
                        grepl("heavy snow|lake-effect snow", unique_events$evtype) == FALSE, 
                        "winter weather", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("fog", unique_events$evtype) == TRUE &
                        grepl("freezing fog", unique_events$evtype) == FALSE,
                        "dense fog", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("high wind|wind", unique_events$evtype) == TRUE &
                        grepl("wind chill|marine|strong wind|thunderstorm wind", unique_events$evtype) == FALSE,
                        "high wind", unique_events$evtype)
unique_events$evtype <- ifelse(grepl("slide", unique_events$evtype) == TRUE,
                               "landslide/mudslide", unique_events$evtype)

unique_events <- unique_events %>%
  group_by(evtype)%>%
  summarise(count = n(), deaths = sum(deaths), injuries = sum(injuries), 
            prop_dmg = sum(prop_dmg), crop_dmg = sum(crop_dmg)) %>%
  arrange(desc(deaths))
```
Now I am down to `r length(unique(unique_events$evtype))` unique events.  The events 
not in official categories were too ambiguous to confidently place into a category. 
Luckily, those event types make up less than 1% of the data.

## Fatality and Injury Data by Event Type

With the hard work out of the way, now I can easily view the results using ggplot2.
```{r human_cost_plot, fig.width = 12, fig.height = 6}
deaths <- unique_events %>% 
  select(evtype, deaths)%>%
  top_n(15, deaths)

p1 <- ggplot(data = deaths, aes(x = reorder(evtype, deaths), y = deaths))+
  geom_point()+
  scale_y_continuous(breaks = seq(0, 6000, by = 500))+
  coord_flip()+
  theme_bw()+
  xlab("Storm Event")+
  ylab("Deaths")+
  ggtitle("Deaths By Event Type")

injuries <- unique_events %>%
  select(evtype, injuries)%>%
  top_n(15, injuries)

p2 <- ggplot(data = injuries, aes(x = reorder(evtype, injuries), y = injuries))+
  geom_point()+
  scale_y_continuous(breaks = seq(0, 92000, by = 10000))+
  coord_flip()+
  theme_bw()+
  xlab("Storm Event")+
  ylab("Injuries")+
  ggtitle("Injuries By Event Type")

grid.arrange(p1, p2, ncol = 2)
```

Tornados are far and away the most harmful events with respect to population health.
In fact, wind related events are heavily represented for both deaths and injuries.

## Economic Impact Data by Event Type
```{r, economic_cost_plot, fig.width = 12, fig.height = 6}
property <- unique_events %>%
  select(evtype, prop_dmg)%>%
  top_n(15, prop_dmg)
p3 <- ggplot(data = property, aes(x = reorder(evtype, prop_dmg), y = prop_dmg/1000000000))+
  scale_y_continuous(breaks = seq(0, 150, by = 10))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  xlab("Storm Event")+
  ylab("Property Damage in Billions")+
  ggtitle("Property Damage By Event Type")

crop <- unique_events %>%
  select(evtype, crop_dmg)%>%
  arrange(desc(crop_dmg))%>%
  top_n(15, crop_dmg)
p4 <- ggplot(data = crop, aes(x = reorder(evtype, crop_dmg), y = crop_dmg/1000000000))+
  scale_y_continuous(breaks = seq(0, 15, by = 1))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  xlab("Storm Event")+
  ylab("Crop Damage in Billions")+
  ggtitle("Crop Damage By Event Type")

grid.arrange(p3, p4, ncol = 2)
```
Hurricanes damage the most property at over 80 billion dollars with tornadoes and storm 
surges both doing over 40 billion dollars in property damage.  Crop damage results make 
intuititve sense with drought and flood events totalling
`r round(sum(crop$crop_dmg[1:2])/sum(unique_events$crop_dmg)*100, digits = 1)`% of the damage.